---
title: "The US Airline Data Analysis Project"
author: "Chang Jia Jie"
date: "2023-03-15"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Load libraries

```{r}
library(dplyr)
library(caTools)
library(ggplot2)
library(tidyr)
library(pROC)
```

Import data

```{r}
df1 <- read.csv('2002.csv')
df2 <- read.csv('2003.csv')
df <- rbind(df1,df2)
airports<-read.csv('airports.csv')
planedata<-read.csv('plane-data.csv')

```

Investigate the data
```{r}
str(df)
summary(df)
```

Handling missing data
```{r}
#Remove six empty columns
df<-df%>%select(-SecurityDelay,-LateAircraftDelay,-CarrierDelay,-WeatherDelay,-NASDelay,-CancellationCode)

#Remove unused columns
df<-df%>%select(-UniqueCarrier, -FlightNum, -TaxiIn, -TaxiOut, -Cancelled, -Diverted)

#Remove rows that contain empty values
df<-df[complete.cases(df),]

summary(df)
```

## Create new x variables

According to the U.S. Department of Transportation, a flight is considered to be a delay if it departs or arrives 15 minutes or more later than the scheduled time. 
```{r}
df$dDelay<-ifelse(df$DepDelay>=15,1,0)
df$aDelay<-ifelse(df$ArrDelay>=15,1,0)
```

Split a day into 4 parts
```{r}
df$time <- ifelse(df$DepTime >= 1800, '1800-2400', 
                        ifelse(df$DepTime >= 1200, '1200-1800',
                               ifelse(df$DepTime >= 600, '0600-1200', '0000-0600')))
```

Create season variable
```{r}
df$date <- ISOdatetime(df$Year, df$Month, df$DayofMonth, 0,0,0,tz="America/Los_Angeles")
spring_2002 <- as.POSIXct("2002-03-20 14:16:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
summer_2002 <- as.POSIXct("2002-06-21 9:24:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
autumn_2002 <- as.POSIXct("2002-09-23 0:56:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
winter_2002 <- as.POSIXct("2002-12-21 20:15:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
spring_2003 <- as.POSIXct("2003-03-20 20:00:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
summer_2003 <- as.POSIXct("2003-06-21 15:10:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
autumn_2003 <- as.POSIXct("2003-09-23 6:47:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
winter_2003 <- as.POSIXct("2003-12-22 2:04:0", format = '%Y-%m-%d %H:%M', tz="America/Los_Angeles")
df$season <- ifelse(df$date >= spring_2002 & df$date < summer_2002, 'Spring',
            ifelse(df$date >= summer_2002 & df$date < autumn_2002, 'Summer',
            ifelse(df$date >= autumn_2002 & df$date < winter_2002, 'Autumn',
            ifelse(df$date >= winter_2002 & df$date < spring_2003, 'Winter',
            ifelse(df$date >= spring_2003 & df$date < summer_2003, 'Spring',
            ifelse(df$date >= summer_2003 & df$date < autumn_2003, 'Summer',
            ifelse(df$date >= autumn_2003 & df$date < winter_2003, 'Autumn', 'Winter')))))))
      
```


## Question 1

Best time of day to fly to minimize delays
```{r}
#compute table of frequency of departure delay by time
freqtime <- table(df$time[df$dDelay == 1])

#compute table of frequency of flight by time
freqtime2 <- table(df$time)

#calculate percentage of departure delay by time
probt <- freqtime/freqtime2

#Plot Bar Chart
barplot(probt, main = "Percentage of Delay by Time of the Day",
        xlab = "Time of the Day", ylab = "Percentage",
        col = "coral")
```

Best day of the week to fly to minimize delays
```{r}
#compute table of frequency of departure delay by day of week
freqd <- table(df$DayOfWeek[df$dDelay == 1])

#compute table of frequency of flight by day of week
freqd2 <- table(df$DayOfWeek)

#calculate percentage of departure delay by day of week
probd <- freqd/freqd2

#Plot Bar Chart
barplot(probd, main = "Percentage of Delay by Day of the Week",
        xlab = "Day of the Week", ylab = "Percentage",
        col = "coral")
```

Best month of the year to fly to minimize delays
```{r}
#compute table of frequency of departure delay by month of year
freqmonth <- table(df$Month[df$dDelay == 1])

#compute table of frequency of flight by month of year
freqmonth2 <- table(df$Month)

#calculate percentage of departure delay by month of year
probm <- freqmonth/freqmonth2

#Plot Bar Chart
barplot(probm, main = "Percentage of Delay Month of the Year",
        xlab = "Month of the Year", ylab = "Percentage",
        col = "coral")
```

Best season of the year to fly to minimize delays
```{r}
#Compute table of frequency of departure delay by season of year
freqseason <- table(df$season[df$dDelay == 1])

#Compute table of frequency of flight by season of year
freqseason2 <- table(df$season)

#Calculate percentage of departure delay by season of year
probs <- freqseason/freqseason2

#Plot Bar Chart
barplot(probs, main = "Percentage of Delay by Season",
        xlab = "Season", ylab = "Percentage",
        col = "coral")
```

## Question 2

Merge planedata and df to create age variable
```{r}
#Rename column 'tailnum' to 'TailNum'
planedata$TailNum <- planedata$tailnum

#Replace blank values with NA
planedata[planedata == ''] <- NA

#Merge df and planedata by 'TailNum'
joindf <- merge(df,planedata, by='TailNum')
str(joindf)

#Remove unused columns
joindf <- joindf%>%select(-TailNum, -tailnum, -type, -manufacturer, -model, -status, -aircraft_type, -engine_type, -year)

#Remove rows with missing values
joindf<-joindf[complete.cases(joindf),]
```

Create Age variable by difference of flying date and issue date 
```{r}
#Change issue_date into POSIXct type
joindf$issue_date <- as.POSIXct(joindf$issue_date, format="%m/%d/%Y", tz="America/Los_Angeles")

#subtract issue date from flying date in terms of days unit
joindf$age <- difftime(joindf$date, joindf$issue_date, units = "days")

#Turn age into numeric variable in terms of years unit
joindf$age <- as.numeric(joindf$age/365)

#Remove rows with NA and negative age values 
joindf <- joindf[joindf$age >= 0, ]
joindf<-joindf[complete.cases(joindf),]
```


Construct correlation test between 'age' and 'arrival delay'
```{r}
#Create a new column, age2 that has rounded age value
joindf$age2 <- round(joindf$age)

#Compute table of frequency of arrival delay by age
freqage <- table(joindf$age2[joindf$aDelay == 1])

#Compute table of frequency of flights by age
freqage2 <- table(joindf$age2)

#Calculate probability of arrival delay by age
probage <- freqage/freqage2

#Turn table into data frame
df3 <- as.data.frame(probage)
names(df3) <- c('Age','Percentage of Delay')

#Turn age variable into integer type
df3$Age <- as.integer(df3$Age)
cor.test(df3$Age,df3$`Percentage of Delay`)
```

Plot bar chart of Delay Percentage vs Age
```{r}
#Plot Bar Chart
barplot(probage, main = "Percentage of Delay by Age",
        xlab = "Age", ylab = "Percentage",
        col = "coral")
```

Create Age Group variable by split age into 3 parts
```{r}
joindf$agegroup <- ifelse(joindf$age2 >= 18, '18-28', 
                        ifelse(joindf$age2 >= 9, '9-17', '0-8'))
```

```{r}
#compute table of frequency of arrival delay by age group
freqgroup <- table(joindf$agegroup[joindf$aDelay == 1])

#compute table of frequency of flight by age group
freqgroup2 <- table(joindf$agegroup)

#calculate percentage of arrival delay by age group
probgroup <- freqgroup/freqgroup2

#Plot Bar Chart
barplot(probgroup, main = "Percentage of Delay by Age Group",
        xlab = "Age Group", ylab = "Percentage",
        col = "coral")
```

## Question 3
```{r}
#Create new variable location that combines Origin and Destination
df1$location <- paste(df1$Origin, df1$Dest, sep='-')
df2$location <- paste(df2$Origin, df2$Dest, sep='-')

#top 10 frequently travel locations in 2002
top_10_values_1 <- sort(table(df1$location), decreasing = TRUE)[1:10]

#frequency of the 10 same locations in 2003
top_10_values_2 <- table(df2$location)[row.names(top_10_values_1)]

# Create data frames for each year's flight frequencies
top_10_values_1 <- as.data.frame(top_10_values_1)
names(top_10_values_1) <- c('location', 'frequency')

top_10_values_2 <- as.data.frame(top_10_values_2)
names(top_10_values_2) <- c('location', 'frequency')

# Combine the data frames into one
flight_data <- rbind(top_10_values_1, top_10_values_2)
flight_data$Year <- rep(c("2002", "2003"), each = 10)

# Compute table
flight_data_wide <- flight_data %>%
  pivot_wider(names_from = Year, values_from = frequency)
flight_data_wide$location2 <- flight_data_wide$location
flight_data_wide <- flight_data_wide %>%
  separate(location2, into = c("Origin", "Dest"), sep = "-")

# Join flight_data_wide with airports based on Origin and Dest columns
flight_data_wide <- flight_data_wide %>%
  left_join(airports, by = c("Origin" = "iata")) %>%
  left_join(airports, by = c("Dest" = "iata"), suffix = c("_origin", "_dest"))

# Replace Origin and Dest columns with airport values
flight_data_wide$Origin <- flight_data_wide$airport_origin
flight_data_wide$Dest <- flight_data_wide$airport_dest

# Remove other columns
flight_data_wide <- flight_data_wide[,1:5]
flight_data_wide
```

Plot line chart
```{r}
line_chart <- ggplot(flight_data, aes(x = location, y = frequency, group = Year, color = Year)) +
  geom_line(aes(y = frequency), linewidth=0.8) +
  labs(title = "Flight Frequencies among Different Locations", x = "Location", y = "Frequency", color = "Year")
line_chart
```

## Question 4
Create detector function
```{r}
detector <- function(Y,M,D,T1,T2,A){
  lvl1 <- df[df$Year==Y & df$Month==M & df$DayofMonth==D & df$CRSArrTime>=T1 & df$CRSArrTime<=T2 & df$Dest==A & df$aDelay==1,] 
  lvl2 <- df[df$Year==Y & df$Month==M & df$DayofMonth==D & df$CRSDepTime>=T1 & df$CRSDepTime<=T2 & df$Origin==A & df$dDelay==1 & df$aDelay==1,] 
  if (nrow(lvl1)>0 & nrow(lvl2>0)){
    cat('There was delay in ', airports$airport[airports$iata==A] ,' airport from time ', T1 , ' to ' , T2, sep = '')
    cat(paste('\nwhich has created cascading delays in', airports$airport[match(lvl2$Dest, airports$iata)], "airport"))
  }
  else if (nrow(lvl1)>0 & nrow(lvl2)==0){
    cat('There was delay in ', airports$airport[airports$iata==A] ,' airport from time ', T1 , ' to ' , T2, ' but it does not create any cascading delay in other airports.' ,sep = '')
  }
  else if (nrow(lvl1)==0){
    cat('There was no delay in ', airports$airport[airports$iata==A] ,' airport from time ', T1 , ' to ' , T2, '. So, it did not cause any cascading delay in other airports.', sep = '')
  }
}
```
example 1:
```{r}
detector(2002,1,31,1500,1600,'CLT')
```

example 2:
```{r}
detector(2003,2,22,2300,2400,'BOS')
```

example 3:
```{r}
detector(2002,1,10,2200,2300,'PHX')
```

## Question 5
```{r}
set.seed(2002)

#refactor variables
joindf[, c('aDelay', "Year", 'Month', 'DayOfWeek', 'dDelay', 'time', 'season')] <- lapply(joindf[, c('aDelay', "Year", 'Month', 'DayOfWeek', 'dDelay', 'time', 'season')], factor)

# perform train test split
train <- sample.split(joindf$aDelay, 0.7)
trainset <- subset(joindf, train == T)
testset <- subset(joindf, train == F)
```

Build logistic regression model to predict flight delay
```{r}
logistic <- glm(aDelay ~ Year+ Month+ DayOfWeek+ AirTime + Distance+ dDelay+ time+  season + age, data = trainset, family = binomial)
summary(logistic)
```

Investigate model prediction result
```{r}
prediction <- predict(logistic, testset, type = 'response')

threshold <- 0.5

Yes.hat <- ifelse(prediction > threshold, 1, 0)

table1 <- table(actual = testset$aDelay, model_prediction = Yes.hat)
table1

table2 <- prop.table(table1,margin = 1)
table2

accuracy <- mean(Yes.hat == (testset$aDelay==1))
accuracy

precision <- table2[1,1]
precision

recall <- table2[2,2]
recall

f1_score <- 2 * (precision * recall) / (precision + recall)
f1_score

roc_obj <- roc(testset$aDelay, Yes.hat)
auc <- auc(roc_obj)
auc
```



